---
title: "*Ceratopteris* ddRADSeq data analysis"
author: "Sylvia Kinosian"
output: html_document
---

All data analysis except the data visualization were run on the University of Utah [Center for High Performance Computing](https://www.chpc.utah.edu/). This or a similar computing cluster is required for this pipeline. Data visualization was performed in R (v. 3.5.2) on Ubuntu 19.04.

# {.tabset}

## Demultiplexing

### STACKS
Tests to see how reads are pairing; not easy in ipyrad

stacks (v. 2.4) process\_radtags pipeline

```{bash eval=FALSE}
process_radtags -P -p /media/kaiser/skinosian/ceratopteris/raw/ -o ./fastqs/ -b barcodes.txt -c -q -r --index_index --renz_1 pstI --renz_2 bfaI
```

#### Parameter Descriptions

**-p**: path to input files
<br>
**-P**, --paired: paired-end reads
<br>
**-o**: path to output files
<br>
**-b**: path to barcodes
<br>
**-c, --clean**: clean data, remove any read with an uncalled base
<br>
**-q, --quality**: discard reads with low quality scores (< 20)
<br>
**-r, -rescue**: rescue barcodes and RAD-Tags
<br>
**--index\_index:** Paired-end with index barcodes on the single and paired-ends
<br>
**--renz\_1**, **--renz\_2:** double-digest restiction enzymes

### PEAR

Paired-end Read Merger (v. 0.9.11)

```{bash eval=FALSE}
pear -f -r -o
```


## ipyrad

```{bash eval=FALSE}
ipyrad -p params-cer1.txt 1234567
```

```{bash eval=FALSE}
------- ipyrad params file (v.0.7.30)-------------------------------------------
cer1                           ## [0] [assembly_name]: Assembly name. Used to name output directories
./                             ## [1] [project_dir]: Project dir (made in curdir if not present)
./rawData/*.fastq.gz           ## [2] [raw_fastq_path]: Location of raw non-demultiplexed fastq files
./all_bc.txt                   ## [3] [barcodes_path]: Location of barcodes file
                               ## [4] [sorted_fastq_path]: Location of demultiplexed/sorted fastq files
denovo                         ## [5] [assembly_method]: Assembly method 
                               ## [6] [reference_sequence]: Location of reference sequence file
pairddrad                      ## [7] [datatype]: Datatype (see docs): rad, gbs, ddrad, etc.
TGCAG, TAG                     ## [8] [restriction_overhang]: Restriction overhang (cut1,) or (cut1, cut2)
5                              ## [9] [max_low_qual_bases]: Max low quality base calls (Q<20) in a read
33                             ## [10] [phred_Qscore_offset]: phred Q score offset (33 is default)
6                              ## [11] [mindepth_statistical]: Min depth for statistical base calling
6                              ## [12] [mindepth_majrule]: Min depth for majority-rule base calling
10000                          ## [13] [maxdepth]: Max cluster depth within samples
0.90                           ## [14] [clust_threshold]: Clustering threshold for de novo assembly
0                              ## [15] [max_barcode_mismatch]: Max number of allowable mismatches in barcodes
2                              ## [16] [filter_adapters]: Filter for adapters/primers (1 or 2=stricter)
35                             ## [17] [filter_min_trim_len]: Min length of reads after adapter trim
2                              ## [18] [max_alleles_consens]: Max alleles per site in consensus sequences
5, 5                           ## [19] [max_Ns_consens]: Max N's (uncalled bases) in consensus (R1, R2)
8, 8                           ## [20] [max_Hs_consens]: Max Hs (heterozygotes) in consensus (R1, R2)
48                             ## [21] [min_samples_locus]: Min # samples per locus for output
20, 20                         ## [22] [max_SNPs_locus]: Max # SNPs per locus (R1, R2)
8, 8                           ## [23] [max_Indels_locus]: Max # of indels per locus (R1, R2)
0.5                            ## [24] [max_shared_Hs_locus]: Max # heterozygous sites per locus (R1, R2)
0, 0, 0, 0                     ## [25] [trim_reads]: Trim raw read edges (R1>, <R1, R2>, <R2) (see docs)
0, 0, 0, 0                     ## [26] [trim_loci]: Trim locus edges (see docs) (R1>, <R1, R2>, <R2)
*                              ## [27] [output_formats]: Output formats (see docs)
                               ## [28] [pop_assign_file]: Path to population assignment file
```

#### Parameter Descriptions

Full documentation for all ipyrad parameters can be found [here](https://ipyrad.readthedocs.io/parameters.html). Below are descriptions of a few key parameters that we chose in our analysis.

[#5](https://ipyrad.readthedocs.io/parameters.html#assembly-method) **Assemby Method** - Since the current version of the *Ceratopteris* genome is relatively low coverage and low quality, we decided to use the *de novo* assembly option for our ddRAD data. ipyrad offers four different assembly methods; for the [denovo method](https://ipyrad.readthedocs.io/methods.html#assembly-methods), raw sequences are assembled without a reference; homology is inferred using [vsearch](https://github.com/torognes/vsearch)  <br>
[#8](https://ipyrad.readthedocs.io/parameters.html#restriction-overhang) **Restriction Overhang** - We used the enzymes PstI and BfaI for our library preparation. [PstI](https://en.wikipedia.org/wiki/PstI) cuts at the 3' end, so you need to reverse compliment of the overhang: TGCAG. [BfaI](https://www.neb.com/products/R0568-BfaI#FAQs%20&%20Troubleshooting) cuts at the 5' end, so is simply: TAG.
<br>
[#16](https://ipyrad.readthedocs.io/parameters.html#filter-adapters) **Filter Adaptors** - We chose the most strict filtering option here, to remove not only barcodes, but Illumina and cutsite adaptors as well. During Step #2, reads are searched for the common Illumina adapter, plus the reverse complement of the second cut site (if present), plus the barcode (if present), and this part of the read is trimmed. 
<br>
[#21](https://ipyrad.readthedocs.io/parameters.html#min-samples-locus) **Min Samples per Locus** - This parameter sets the minimum number of samples that must have data for a given locus in the final data output. We chose to go with a relatively high number of minimum samples - 48 or half of the total samples. We wanted our final output file to have data from a at least half (if not a majority) of our individuals.

#### Analysis of raw reads & loci

```{r eval=TRUE}
d <- read.csv("./structure/reads.raw.txt", sep = ',', header = T)

hist(d[,2], main = "raw reads", xlab = "number of reads", breaks = 20)
hist(d[,3], main = "number of loci in assembly", xlab = "number of loci", breaks = 20)
```

## Population genetic analysis

#### STRUCTURE

The population genetics analysis program STRUCTURE (v. 2.3.4) was used to identify populations within the genus. For each k value, we ran 50 replicates (NOTE: include params files in github).

We used the [Cluster Markov Packager Across K (CLUMMPAK)](http://clumpak.tau.ac.il/) to combine replicates within k values.

Visualization of STRUCTURE was done in R with a custom script.

```{r eval=FALSE}
# load in ks from clummpack files
k2 <- read.csv("k2.txt", sep = '', header = F)
k2 <- k2[,-(1:5)]
k3 <- read.csv("k3.txt", sep = '', header = F)
k3 <- k3[,-(1:5)]
k4 <- read.csv("k4.txt", sep = '', header = F)
k4 <- k4[,-(1:5)]
k5 <- read.csv("k5.txt", sep = '', header = F)
k5 <- k5[,-(1:5)]
k6 <- read.csv("k6.txt", sep = '', header = F)
k6 <- k6[,-(1:5)]
k7 <- read.csv("k7.txt", sep = '', header = F)
k7 <- k7[,-(1:5)]

# names file includes individual ids, species names, and geographic locations
names <- read.csv("names.csv", sep = ',', header = T)

x <- as.data.frame(matrix(ncol = 18, nrow = 60))
x[,1:2] <- k2
x[,3:5] <- k3
x[,6:9] <- k4
x[,10:14] <- k5
x[,15:20] <- k6
x[,21:27] <- k7
x[,28:31] <- names

klist <- list(x[,1:2], x[,3:5], x[,6:9], x[,10:14], x[15:20], x[,21:27])

structure_plot(names, ninds = 67, klist)

#######################################################
# functions needed
######################################################

# plotting and labeling function
structure_plot <- function(labels, ninds = 60, klist){
	# define colors
	cols <- c('#A8FFFD', '#A39D9D','#FFFF00', '#ff5a5a', '#69C261', '#26CDCD', '#B862D3','#C1C6FF')
	# unique label names
	sp.names <- as.character(unique(labels))
	#n <- as.data.frame(matrix(ncol = 1, nrow = ninds))
	#n[,1] <- names
	# locations of each column
	b <- as.data.frame(matrix(ncol = 1, nrow = ninds))
	b[,1] <- barplot(t(k6list[[1]][1]), beside= F, col= cols, cex.name= 1, cex.axis= 1.2, border = 1, space = 0.05, xaxt = 'n', yaxt = 'n', cex.lab = 1, cex.main = 2)
	# find locations for labels in the barplot
	my.mean <- tapply(X = b[,1], INDEX = labels, mean)
	my.min <- tapply(X = b[,1], INDEX = labels, min)
	my.max <- tapply(X = b[,1], INDEX = labels, max)
	# data frame for plotting
	d <- sp_labels(names = sp.names, min = my.min, mean = my.mean, max = my.max)
	# plot
	plot_q_per_chain(klist)
	text(cex = 1.5, x = (d[,2]-0.3), y = -0.7, labels = d[,1], xpd=NA, srt=50, font=3)
	# lines
	for (i in 1:length(d[,1])){
		lines(x = d[i,3:4] , y = rep(-0.1, 2), lwd = 2.5, col = "black", xpd = NA)
	}

}

# create labels
sp_labels <- function(names, min, mean, max, ...){
	d <- as.data.frame(matrix(nrow = length(names), ncol = 4))
	for (j in 1:length(names)){
			d[j,1] <- names[j]
			d[j,3] <- min[[j]][1]
			d[j,2] <- mean[[j]][1]
			d[j,4] <- max[[j]][1]
	}
	return(d)
}

# plot chains with species and geography labels 
plot_q_per_chain <- function(kqlist, ...){
	cols <- c('#A8FFFD','#B862D3', '#A39D9D','#FFFF00', '#69C261', '#FF59AC', '#26CDCD',  '#C1C6FF') 	
	par(mfrow = c(length(kqlist),1), mar = c(1,3,3,1) + 0.1, oma = c(15,0,0,0), mgp = c(1,1,0))
	chain <- seq(1, length(kqlist), 1) 
	for(i in 1:length(kqlist)){
		barplot(t(kqlist[[i]]), beside= F, col= cols, border = 1, space = 0.05, xaxt = 'n', yaxt = 'n', main = paste("k =", chain[i]+1, sep = ' '), cex.lab = 1.2, cex.main = 1.6)
		# y axis
		axis(2, at = c(0, 0.25, 0.5, 0.75, 1), cex.axis = 1, las = 2, pos = -0.2)
 	}
	# x axis, rotating labels
	# species labels
		# location labels (make new df for this one?)
	#text(cex = 1.7, x = (e[,3]-4), y = -0.9, labels = e[,1], xpd=na, srt=50, font=3)

}
structure_plot(labels = x[,31], klist = klist)

############################################################################
############################################################################
# for k=5 only
plot_k5 <- function(kqlist, ...){
	cols <- c('#A8FFFD','#B862D3', '#A39D9D','#FFFF00', '#69C261', '#FF59AC', '#26CDCD',  '#C1C6FF') 	
	#cols <- c('#A8FFFD', '#B862D3','#A39D9D','#FFFF00', '#ff5a5a', '#69C261', '#26CDCD', '#C1C6FF') 	
	par(mfrow = c(length(kqlist),1), mar = c(1,3,3,1) + 0.1, oma = c(15,0,0,0), mgp = c(1,1,0))
	chain <- seq(1, length(kqlist), 1) 
	for(i in 1:length(kqlist)){
		barplot(t(kqlist[[i]]), beside= F, col= cols, border = 1, space = 0.05, xaxt = 'n', yaxt = 'n', main = substitute(paste(italic(K)," = 6")), cex.lab = 1.6, cex.main = 2.0, ylab = "Proportion of genome")
		# y axis
		axis(2, at = c(0, 0.25, 0.5, 0.75, 1), cex.axis = 1, las = 2, pos = -0.2)
 	}
}

plot_k5(k6list)

# x axis, rotating labels
text(cex = 1.7, x = (d[,2]-0.4), y = -0.3, labels = d[,1], xpd=NA, srt=50, font=3)

for (i in 1:length(d[,1])){
	lines(x = d[i,3:4] , y = rep(-0.05, 2), lwd = 2.5, col = "black", xpd = NA)
}

```

#### Tetrad

ipyrad inferring quartet trees with [tetrad](https://nbviewer.jupyter.org/github/dereneaton/ipyrad/blob/master/tests/cookbook-quartet-species-tree.ipynb)

Command line, on a server
```{bash eval=FALSE}
# start an interactive job
srun -t 24:00:00 -n 24 -N 1 -A usubio-kp -p usubio-kp --pty /bin/bash -l

# start ipcluster
ipcluster start --n=24 --daemonize
```

Python
```{perl eval=FALSE}
import ipyrad.analysis as ipa
import ipyparallel as ipp
import toytree

ipyclient = ipp.Client()
print("connected to {} cores".format(len(ipyclient)))
# connected to 24 cores

tet = ipa.tetrad(
	name="cer",
	seqfile="./final.snps.phy",
	mapfile="./final.snps.map",
	nboots=50,
	)

# loading seq array [48 taxa x 16866 bp]
# max unlinked SNPs per quartet (nloci): 4491

tet.run(ipyclient=ipyclient)
# host comput node: [24 cores] on kp382
# inferring 3321960 induced quartet trees
# [####################] 100%  initial tree | 0:00:33 |
# [####################] 100%  boot 100     | 0:47:22 |
```

Plotting
```{r eval=FALSE}
library(ape)
library(phytools)
# https://github.com/willpearse/willeerd
library(willeerd)

t <- read.tree(file.choose("cer.tree"))
# Enter file name: cer.tree

t
# Phylogenetic tree with 48 tips and 47 internal nodes.

#Tip labels:
#	Pt03_R2_, Pt04_R2_, Pt02_R1_, Pt05_R1_, Pt06_R2_, Pt01_R2_, ...

#Rooted; no branch lengths.

t <- root(phy=t, outgroup="AcAu", resolve.root=T)
plot(t, cex = 0.8)

t$tip.label <- as.character(cnames$str_sp)

############################################################
# plot with species names

plot(c, direction = "up", show.tip.label = FALSE)

# thalictroides
willeerd.tiplabels(tip = c(1:17, 19, 35:37, 39:61), pch = 21, col = "#000000", bg = "#000075", cex = 2)

# sp.
willeerd.tiplabels(tip = c(34, 54), pch = 21, col = '#000000', bg = "#AAFFC3", cex = 2)

# cornuta
willeerd.tiplabels(tip = c(29, 30:32), pch = 21, col = '#000000', bg = "#F58231", cex = 2)

# pteridoides
willeerd.tiplabels(tip = c(22, 24:28), pch = 21, col = '#000000', bg = "#FFE119", cex = 2)

# gaudichaudii
willeerd.tiplabels(tip = c(18, 23), pch = 21, col = '#000000', bg = "#E6194B", cex = 2)

# richardii
willeerd.tiplabels(tip = c(20, 21, 38), pch = 21, col = '#000000', bg = "#3CB44B", cex = 2)

# Acrostichum
willeerd.tiplabels(tip = 33, pch = 21, col = '#000000', bg = "#800000", cex = 2)

legend(1.5, 250, legend = c("C. gaudichaudii", "C. cornuta", "C. pteridoides", "C. richardii", "C. sp.", "C. thalictroides", "A. aureum"), pch = 19, cex = 2, col = c("#E6194B", "#F58231", "#FFE119", "#3CB44B", "#AAFFC3", "#000075", "#800000"), bty= "n", text.font = 3)

# plot with geography

plot(c, direction = "up", show.tip.label = FALSE)

# Australia
willeerd.tiplabels(tip = c(15, 33, 39, 40:52), pch = 21,  col = '#000000', bg ='#000000', cex = 1.2)
# #FFFF00

# Central America
willeerd.tiplabels(tip = c(2,5,20,21,56:58), pch = 21, col = '#000000', bg ='#4363D8', cex = 1.2)

# North America
willeerd.tiplabels(tip = c(4,6:8, 10:12, 17, 25:28, 35), pch = 21, col = '#000000', bg ='#911EB4', cex = 1.2)

# South America
willeerd.tiplabels(tip = c(9, 22, 24, 34, 36:38), pch = 21, col = '#000000', bg ='#F032E6', cex = 1.2)

# Asia
willeerd.tiplabels(tip = c(1,3,11,13,14,16,18,19,23,53:55,59,60,61), pch = 21, col = '#000000', bg ='#A9A9A9', cex = 1.2)

# Africa
willeerd.tiplabels(tip = c(29:32), pch = 21, col = '#000000', bg = '#42D4F4', cex = 1.2)

legend(1.5, 250, legend = c("Australia", "Central America", "North America", "South America", "Asia", "Africa"), pch = 19, cex = 1.5, col = c("#000000", "#4363D8", "#911EB4", "#F032E6", "#A9A9A9", "#42D4F4"))

# plot with new names, derived from structure

plot(t, direction = 'up', show.tip.label = FALSE)

# thalictroides 1
willeerd.tiplabels(tip = c(19, 4:10, 54), pch = 21, col = '#000000', bg = "#AAFFC3", cex = 2)

# thalictroides 2
willeerd.tiplabels(tip = c(1, 2, 36, 37, 39:53, 55:61), pch = 21, col = "#000000", bg = "#000075", cex = 2)

# cornuta
willeerd.tiplabels(tip = c(3, 22, 29:32, 34), pch = 21, col = '#000000', bg = "#F58231", cex = 2)

# pteridoides
willeerd.tiplabels(tip = c(24:28), pch = 21, col = '#000000', bg = "#FFE119", cex = 2)

# gaudichaudii
willeerd.tiplabels(tip = c(11:18, 23, 35), pch = 21, col = '#000000', bg = "#E6194B", cex = 2)

# richardii
willeerd.tiplabels(tip = c(20, 21, 38), pch = 21, col = '#000000', bg = "#3CB44B", cex = 2)

# Acrostichum
willeerd.tiplabels(tip = 33, pch = 21, col = '#000000', bg = "#800000", cex = 2)

legend(1.5, 250, legend = c("C. gaudichaudii", "C. cornuta", "C. pteridoides", "C. richardii", "C. thalictroides 1", "C. thalictroides 2", "A. aureum"), pch = 19, cex = 2, col = c("#E6194B", "#F58231", "#FFE119", "#3CB44B", "#AAFFC3", "#000075", "#800000"), bty= "n", text.font = 3)


#############################################################
# trying to sort one column by another with the same names in a different order
# m <- match(names[,1], c_names[,1])
# match_names <- match_names[order(m),]
```

#### gbs2ploidy

I converted the VCF file output by ipyrad to the gbs2ploidy format with this [Python script](https://github.com/carol-rowe666/vcf2hetAlleleDepth) written by Carol Rowe.

```{r eval=FALSE}
library(gbs2ploidy)

ids <- read.csv("./HAD_ID.csv")
het <- as.matrix(read.table("./hetAlleleDepth.txt", header = F))

a <- seq(1, 142, 2)
b <- seq(2, 142, 2)

cov1 <- het[,a]
cov2 <- het[,b]

propOut <- estprops(cov1 = cov1, cov2 = cov2, props = c(0.25, 0.33, 0.5, 0.66, 0.75), mcmc.nchain = 3, mcmc.steps = 10000, mcmc.burnin = 1000, mcmc.thin = 5)

dev.new()
par(mfrow = c(3, 3))

for(i in 1:71){
  plot( propOut[[i]][,3], ylim = c(0,1), axes = FALSE, xlab = "ratios", ylab = "proportions")
  axis( 1, at = 1:5, c("1:3", "1:2", "1:1", "2:1", "3:1"))
  axis(2)
  box()
  segments(1:5, propOut[[i]][,1], 1:5, propOut[[i]][,5])
  title(main = paste("sample name =", ids[[1]][i]))
}

dev.off()
```

#### Map

```{r eval=FALSE}
library(fields)

map('world', fill = T, col = 'white', border = 'white', bg = 'lightgrey')

# cornuta
points(p[2:5,5], p[2:5,4], col = "#000000", bg = "#F58231", cex = 2, pch = 21)

# gaudichaudii
points(p[6:7,5], p[6:7,4], col = "#000000", bg = "#E6194B", cex = 2, pch = 21)

#pteridoides
points(p[8:13,5], p[8:13,4], col = "#000000", bg = "#FFE119", cex = 2, pch = 21)

# richardii
points(p[14:16,5], p[14:16,4], col = "#000000", bg = "#3CB44B", cex = 2, pch = 21)

# thalictroides
points(p[17:39,5], p[17:39,4], col = "#000000", bg = "#000075", cex = 2, pch = 21)

legend(-170, -2, legend = c("C. gaudichaudii", "C. cornuta", "C. pteridoides", "C. richardii", "C. thalictroides"), bg = "white", pch = 19, cex = 1.5, col = c("#E6194B", "#F58231", "#FFE119", "#3CB44B", "#000075"), text.font = 3)
```

## Program Versions

Below is a list of all program versions used in this analysis. **Please note** that newer versions of these software packages *may* work for this pipeline, but be aware that usage often changes with new verions. 

[ipyrad](https://ipyrad.readthedocs.io/) [release: 0.7.30](https://github.com/dereneaton/ipyrad/releases/tag/0.7.30)

[STRUCTURE v.2.3.4](https://web.stanford.edu/group/pritchardlab/structure_software/release_versions/v2.3.4/html/structure.html)

[Cluster Markov Packager Across K (CLUMMPAK)](http://clumpak.tau.ac.il/) 

[Perl 5](https://www.perl.org/)

[Python 2.7.13](https://www.python.org/downloads/release/python-2713/)

[R v. 3.5.2](https://www.r-project.org/)

